Demographic inference using the SFS with \moments and \demes

### 2025-08-26 chapter/paper.tex
Update text, revise figure 1

\emph{Keywords:} Demographic inference, site frequency spectrum, \aprcomment{any others?}

% TODO:
% - orcid numbers if any - done
% - interline double, also for references - done
% - line numbers - done
% - keywords - done
% - running head - done
% - Figures in separate files (named Figure 1, Figure2, etc) for the final submission, legends listed at the end of the chapter. - done

@@

This chapter focuses on multi-population demographic inference from the SFS
using \moments \cite{jouganous2017inferring}.

Briefly, \moments \ldots.

\reviewercomment{A short paragraph
    briefly describing the principle of the moments methods would be great. How
    does it differ from other sfs based methods? Just for the readers to get an
idea, then directing them to the relevent literature for more details.}

@@

Other evolutionary mechanisms are known to affect allele frequency dynamics and
the SFS. This presents a challenge for demographic inference, as the observed
SFS may be distorted by non-demographic processes. However, this also presents
an opportunity to learn about different evolutionary processes, if demographic
history can be controlled for. The SFS has been used to infer the distribution
of fitness effects of new mutations, primarily for nonsynonymous variation in
coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
kim2017inference}); to scan for historical selective sweeps
(e.g., \cite{kim2002detecting, nielsen2005genomic}); and to understand how
selection on quantitative traits impacts the genetic architecture underlying
those traits (e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS
has also been used to infer relative mutation rates and how they have changed
over time \cite{dewitt2021nonparametric}. Each of these analyses typically
requires partitioning the data in some way -- by functional annotation,
mutation context, local recombination rate, etc -- so that comparisons can be
made across SFS from different classes of mutations.

@@

\reviewercomment{Would it be possible to include some information regarding possible local optima? Do you recommend starting from different initial conditions?}

@@

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize()}
function. The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. Any value in the
demographic model that is not specified in the parameters file will remain
unchanged, and is therefore treated as a fixed parameter. Here, we also assume
we have an estimate for the total mutation rate, $U$.

@@

\reviewercomment{Mirroring my comments in the introduction, I would suggest to add a few sentences to highlight how moments differ from other approaches, in terms of caveats and advantages. No need to go in details, just the take-home message and references for further details.}

@@

\begin{figure}
    %\includegraphics{../example1/fig1.pdf}
    \caption{
        Inference using simulated data. (A) The true demographic model,
        including an expansion in the ancestral population and symmetric
        migration between the descendent populations. (B--D) The simulated
        data, showing both marginal and the joint SFS. (E--G) Residuals
        between the data SFS (in D) and the predicted SFS from the inferred
        demographic models in (H--J), resp.  The two misspecified models (H and
        I) provide worse fits to the data and both overestimate the split time
        due to disallowing any size change in the ancestral population.
        \reviewercomment{it might be good to label the colorbars for the joint spectra as 'allele frequency' for people that are unfamiliar with what those plots are showing}
    }
    \label{fig:im}
\end{figure}

>>>>

\emph{Keywords:} Demographic inference, site frequency spectrum, diffusion
equation, moments equations

@@

@@

This chapter focuses on multi-population demographic inference from the SFS
using \moments \cite{jouganous2017inferring}. Briefly, \moments computes the
expected SFS by numerically solving a system of ordinary differential equations
closely related to the diffusion approximation \cite{gutenkunst2009inferring}.
The flexibility of the numerical methods in \moments allows for non-constant
population size histories including up to five populations at a time, with gene
flow. Selection can also be incorporated, allowing for recessive, additive or
dominant selection, or over- or underdominance at a locus. This differs from
most coalescent-based approaches (e.g.,
\cite{excoffier2011fastsimcoal,dewitt2021nonparametric,dilber2024faster}),
which are typically limited to neutrality.

Many evolutionary mechanisms are known to affect allele frequency dynamics and
the SFS. This presents a challenge for demographic inference, as the observed
SFS may be distorted by non-demographic processes. However, this also presents
an opportunity to learn about different evolutionary processes, if demographic
history can be controlled for. The SFS has been used to infer the distribution
of fitness effects of new mutations, primarily for nonsynonymous variation in
coding regions (e.g., \cite{eyre2006distribution, boyko2008assessing,
kim2017inference}); to scan for historical selective sweeps
(e.g., \cite{kim2002detecting, nielsen2005genomic}); and to understand how
selection on quantitative traits impacts the genetic architecture underlying
those traits (e.g., \cite{patel2024conditional, ragsdale2024archaic}). The SFS
has also been used to infer relative mutation rates and how they have changed
over time \cite{dewitt2021nonparametric}. Each of these analyses typically
requires partitioning the data in some way -- by functional annotation,
mutation context, local recombination rate, etc -- so that comparisons can be
made across SFS from different classes of mutations.

@@

@@

Taking the demographic model, parameters file, and observed SFS together, we
can perform inference using the \texttt{moments.Demes.Inference.optimize()}
function. The initial parameter guesses are given by the input demographic
model, which may be perturbed using a keyword argument. The likelihood surface
may be rugged or flat in some dimensions, and optimization routines sometimes
fail to converge to the global (or sometimes even local) optimum. Thus, we
often try many different initial guesses for parameter values, which can be
randomly assigned using the \texttt{perturb} keyword argument. Running multiple
iterations of optimization with different optimization functions is also
recommended to ensure convergence.

Any value in the demographic model that is not specified in the parameters file
will remain unchanged, and is therefore treated as a fixed parameter. Here, we
also assume we have an estimate for the total mutation rate, $U$.

@@

@@

\begin{figure}
    %\includegraphics{../example1/fig1.pdf}
    \caption{
        Inference using simulated data. (A) The true demographic model,
        including an expansion in the ancestral population and symmetric
        migration between the descendent populations. (B--D) The simulated
        data, showing both marginal and the joint SFS. (E--G) Residuals
        between the data SFS (in D) and the predicted SFS from the inferred
        demographic models in (H--J), resp.  The two misspecified models (H and
        I) provide worse fits to the data and both overestimate the split time
        due to disallowing any size change in the ancestral population.
    }
    \label{fig:im}
\end{figure}
